---
title: "Analysis of the most harmful and lethal products"
author: "Akvilė Višniauskaitė ir Ieva Pudžiuvelytė (AI)"
date: '2022-02-17'
output: pdf_document
---

# Description of the task

A public institution that regulates sales of the products is willing to label 
dangerous products accordingly. The information about the products that cause 
the most serious symptoms was requested. More formally, answers to the following 
questions needed to be provided:

1. Top 5 most serious (the most harmful to health) symptoms
2. Which industry products cause death most frequently?
3. How many and what kind of products should be banned in order to reduce number
   of deaths by 20%?
4. What other interesting insight can be found?
5. Additional library to visualise data 
6. At least five additional questions:

- What products were concomitant to cause death?
- Were there any other reports (not lethal) of folligen solution therapy?
- What are other symptoms caused by "Exemption 4"?
- Which gender uses product "Exemption 4" more?
- Which age group uses "Exemption 4" the most?

## Comments about report

The first, second, third and additional questions are written down in the report
in *italics* for clarity. The additional library for data visualisation was 
described in a separate section named "Visualisation of proportions".

\newpage

```{r echo=FALSE}
#install.packages('treemap')
#install.packages('stringr')
#install.packages('FSA')
library(FSA)
```

```{r echo=FALSE}
data <- readRDS("data.rds")
tm <- readRDS("tm.rds")
```

```{r echo=FALSE}
library(treemap)
# Color vector
pastel_pie_color_vector <- c('#D6C9BD', '#FFDCCD', '#F7F6E6')
```

# Description of data

Data was taken from a database CFSAN Adverse Event Reporting System (CAERS),
which was collected in the period of January 2018 to December 2020 [1, 2, 3]. 
The database consists of compliants by healthcare professionals and consumers 
from the United States and are related to foods, dietary supplements, or 
cosmetics.

The original data for analysis is presented in a form of a comma-separated value 
table. The original dataset [2] had few modifications for data consistency. 
Consequentially, the dataset of reports, which was used for analysis, 
consisted of 10 columns with the following information:

- id - an identifier of the report
- created - a date when the report was created
- date - a date when symptoms occurred
- type - a label whether the product is considered as "suspect" or "concomitant"
- product - a name of the reported product
- code - a numeral identifier of an industry of the reported product
- industry - a literal identifier of an industry of the reported product
- age - age of a patient (years)
- sex - gender of a patient ("M" - male, "F" - female, "U" - unidentified)
- outcome - symptoms that the product caused (or was suspected to cause) 

# Harmful products

*Which are top 5 most serious (the most harmful to health) symptoms?*

After the data analysis, it was observed that there were cases of death 
reported, which could naturally be included as the most harmful symptom to one's 
health. However, the analysis regarding reported cases of death will be 
presented below, thus this section is an attempt to filter the most serious 
symptoms besides death of an individual. 

Seriousness can be interpreted differently. In our view, any form of stroke, 
cancer, paralysis, or anything related to brain or heart are serious symptoms, 
or, more precisely, outcomes that appeared after consuming or using the product. 
Therefore, we extracted and listed out these groups, although, the symptoms 
within these groups are rather diverse.

There were reports of different types:

- death (including sudden, foethal, cardiac, 
  and brain): 9
- stroke (including ischaemic, haemorrhagic, 
  and brain stem): 5
- cancer (including stage IV and metastatic
  forms): 70
- related to brain: 15
- related to heart: 8
- paralysis: 4

In order to give a clear answer to the question, we picked out five most serious
symptoms from these groups that were apparent in the dataset of the reports:

1. Brain death
2. Ischaemic stroke
3. Haemorrhagic stroke
4. Metastatic cancer forms
5. Stage IV cancer forms

```{r include=FALSE}
colnames(tm)[grep('death', colnames(tm))]
colnames(tm)[grep('stroke', colnames(tm))]
colnames(tm)[grep('cancer', colnames(tm))]
colnames(tm)[grep('brain', colnames(tm))]
colnames(tm)[grep('heart', colnames(tm))]
colnames(tm)[grep('paralysis', colnames(tm))]
```

\  

# Lethal products

```{r include=FALSE}
# Saving a subset of reports that included death as an outcome:
lethal_reports <- data[grep('death', data$outcome),]

# Overall number of lethal outcomes:
length(lethal_reports$outcome)

# 20% of all:
length(lethal_reports$outcome)*0.2

# Number of different outcomes with death included:
length(unique(lethal_reports$outcome))

# Number of different products that caused death:
length(unique(lethal_reports$product))

# Number of lethal reports with type "suspect":  
length(lethal_reports[lethal_reports$type == "suspect", c("product")])

# Number of lethal reports with type "concomitant":
length(lethal_reports[lethal_reports$type == "concomitant", c("product")])

# The concomitant case
lethal_reports[lethal_reports$type == "concomitant", ]
```


The overall number of lethal outcomes was `r length(lethal_reports$outcome)`.
These outcomes could be grouped to `r length(unique(lethal_reports$outcome))` 
different types of outcomes that included death as one of them.

\  

*What products were concomitant to cause death?*

```{r include=FALSE}
data[grep(lethal_reports[lethal_reports$type == "concomitant", c("product")], 
          data$product),]
```

Only 
`r length(lethal_reports[lethal_reports$type == "concomitant", c("product")])` 
of all reports had a type **concomitant**, the remaining ones were only 
suspected to cause death. The concomitant case was about 
`r lethal_reports[lethal_reports$type == "concomitant", c("product")]` from 
`r lethal_reports[lethal_reports$type == "concomitant", c("industry")]` 
industry that caused death for a male patient, who was
`r lethal_reports[lethal_reports$type == "concomitant", c("age")]` years old.

\  

*Were there any other reports of folligen solution therapy?*
 
```{r include=FALSE}
data[grep("folligen solution therapy", data$product),]
```

The concomitant report regarding the death of the 29 years old male was the only
one with `r lethal_reports[lethal_reports$type == "concomitant", c("product")]`
included.

\  

## Reducing number of deaths

```{r include=FALSE}
# Frequency of each product that caused death:
product_freq <- table(lethal_reports$product)
product_freq <- product_freq[order(product_freq, decreasing = TRUE)]
max(table(lethal_reports$product))
sum(table(lethal_reports$product))

most_lethal_product_name <- row.names(product_freq)[1]
most_lethal_product_name_cap <- gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2",   
                                      most_lethal_product_name,
                                      perl = TRUE)
most_lethal_product_industry <- lethal_reports[
                                data$product == most_lethal_product_name, 
                                c("industry")][1]

most_lethal_number_deaths <- max(table(lethal_reports$product))
most_lethal_percen_deaths <- max(table(lethal_reports$product))/
                                sum(table(lethal_reports$product))*100

```

*How many and what kind of products should be banned in order to reduce number of deaths by 20%?*

In order to answer the question, which product group should be banned to reduce
the number of deaths by 20%, we would have to find the product or several 
products that makes up 20% of all deaths reported. 

- Number of all death cases: `r length(lethal_reports$outcome)`
- Number of death cases that makes up 20%: `r 0.2*length(lethal_reports$outcome)`

Formally, the goal is to find a group of products that would be the suspected
cause of at least `r 0.2*length(lethal_reports$outcome)` death cases.

However, after performed analysis of the data, such group cannot be found, 
since there is one product named "`r most_lethal_product_name_cap`" from 
`r most_lethal_product_industry` industry which is responsible for 
`r most_lethal_number_deaths` death cases, which makes
up `r round(most_lethal_percen_deaths, 2)`%
of all cases. Therefore, there are two possible subsequent actions:

- Ban all other products except "`r most_lethal_product_name_cap`" - this action 
  would decrease the number of deaths by 
  `r round(100 - most_lethal_percen_deaths, 2)`%

- Ban "`r most_lethal_product_name_cap`" - this action would reduce the 
  number of deaths by `r round(most_lethal_percen_deaths, 2)`%

Therefore, if the goal is to reduce the number of deaths by at least 20%, 
"`r most_lethal_product_name_cap`" should be prohibited.

\  

*Which industry products cause death most frequently?*

Based on the analysis described above, the industry whose products cause death 
most frequently is `r most_lethal_product_industry`, since this is the 
industry to which "`r most_lethal_product_name_cap`" belongs.

\  

## Visualisation of proportions

Below a pie chart presenting the proportions of the death causing products is
shown. For more clarity, "treemap" package was used as an alternative to 
visualise proportions, since the most common argument against pie charts is that 
varied angles complicate the process of detecting differences between displayed
areas [4].

```{r echo=FALSE, fig.height=6, fig.width=6, out.width="50%" }
# Summing up frequency of other products
other <- sum(product_freq[3:length(product_freq)])

# Setting labels for a pie chart
label_1 <- paste(round((product_freq[1]/length(lethal_reports$outcome))*100, 2), 
                 '%')

label_2 <- paste(round((product_freq[2]/length(lethal_reports$outcome))*100, 2), 
                 '%')

label_other <- paste(round((sum(product_freq[3:length(product_freq)])
                            /length(lethal_reports$outcome))*100, 2), '%')

# Drawing the pie chart
pie(c(product_freq[1], product_freq[2], other), 
    labels=c(label_1, label_2, label_other), 
    main="", 
    col=pastel_pie_color_vector)

# Setting labels for a treemap
label_1 <- gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2",   
                                      row.names(product_freq)[1],
                                      perl = TRUE)

label_2 <- gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2",   
                                      row.names(product_freq)[2],
                                      perl = TRUE)

name <- c(label_1, label_2, 'Other products')
frequency <- c(product_freq[1], product_freq[2], other)

df <- data.frame(name, frequency)

# Drawing the treemap
p <- treemap(df, index=c('name'), vSize='frequency', 
                             palette=pastel_pie_color_vector, 
             title='',
             fontsize.labels=c(16, 10, 10), fontsize.title=16, 
             align.labels=list(c("center", "center")))

```

Fig. 1. The proportions of products that were reported with death as an outcome

\  

*What are other outcomes caused by "`r most_lethal_product_name_cap`"?*

```{r echo=FALSE, include=FALSE}
library(stringr)
exemption_4_data <- data[grep(most_lethal_product_name, data$product), ]
exemption_4_data <- na.omit(exemption_4_data)
exemption_4_outcomes <- c(str_split(exemption_4_data$outcome, ', '))
exemption_4_outcomes <- unlist(exemption_4_outcomes)
unique(exemption_4_outcomes)
```

There were 10 other outcomes listed for "`r most_lethal_product_name_cap`":

1. hospitalization
2. medically important
3. patient visited healthcare provider
4. other outcome
5. other seriousness
6. patient visited ER
7. disability
8. life threatening
9. required intervention
10. congenital anomaly

\  

*Which gender uses the product "`r most_lethal_product_name_cap`" more?*

```{r echo=FALSE}
exemption_4_gender_freq <- table(exemption_4_data$sex)
```

The gender group that uses "`r most_lethal_product_name_cap`" more is the female 
group, which makes up 
`r round((exemption_4_gender_freq[1]/length(exemption_4_data$outcome))*100, 2)`%
of all reports of this product.

```{r echo=FALSE, figures-side, fig.height=6, fig.width=6, out.width="50%" }
label_1 <- paste(round((exemption_4_gender_freq[1]/length(exemption_4_data$outcome))*100, 2), 
                 '%')

label_2 <- paste(round((exemption_4_gender_freq[2]/length(exemption_4_data$outcome))*100, 2), 
                 '%')

label_3 <- paste(round((exemption_4_gender_freq[3]/length(exemption_4_data$outcome))*100, 2), 
                 '%')

# Drawing the pie chart
pie(c(exemption_4_gender_freq[1], exemption_4_gender_freq[2], 
      exemption_4_gender_freq[3]), 
    labels=c(label_1, label_2, label_3), 
    main="", 
    col=pastel_pie_color_vector)

# Setting labels for a treemap
label_1 <- gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2",   
                                      row.names(exemption_4_gender_freq)[1],
                                      perl=TRUE)

label_2 <- gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2",   
                                      row.names(exemption_4_gender_freq)[2],
                                      perl=TRUE)

label_3 <- gsub("(^|[[:space:]])([[:alpha:]])", "\\1\\U\\2",   
                                      row.names(exemption_4_gender_freq)[3],
                                      perl=TRUE)

name <- c(label_1, label_2, label_3)
frequency <- c(exemption_4_gender_freq[1], exemption_4_gender_freq[2], 
               exemption_4_gender_freq[3])

df <- data.frame(name, frequency)

# Drawing the treemap
p <- treemap(df, index=c('name'), vSize='frequency', 
                             palette=pastel_pie_color_vector, 
             title='',
             fontsize.labels=c(16, 10, 10), fontsize.title=16, 
             align.labels=list(c("center", "center")))
```
Fig. 2. The proportions of gender groups that used product "Exemption 4"

\  

*Which age group uses "`r most_lethal_product_name_cap`" the most?*

```{r echo=FALSE, include=FALSE}
exemption_4_age_freq <- table(exemption_4_data$age)

exemption_4_age_freq_df <- as.data.frame(exemption_4_age_freq)
exemption_4_age_freq_df[exemption_4_age_freq_df$Freq == 
                          max(exemption_4_age_freq_df$Freq),]
```

```{r echo=FALSE, include=FALSE}
exemption_4_data[exemption_4_data$ age == 60,c('sex')]
length(exemption_4_data[exemption_4_data$ age == 60,c('sex')])
sum(exemption_4_data[exemption_4_data$ age == 60,c('sex')]=='F')
sum(exemption_4_data[exemption_4_data$ age == 60,c('sex')]=='M')
sum(exemption_4_data[exemption_4_data$ age == 60,c('sex')]=='U')
```

"`r most_lethal_product_name_cap`" is mostly reported by 60 year old people, of 
which majority are female 
(`r sum(exemption_4_data[exemption_4_data$ age == 60,c('sex')]=='F')` out of 
`r length(exemption_4_data[exemption_4_data$ age == 60,c('sex')])`).

```{r echo=FALSE}
histFromSum(exemption_4_age_freq, 
            xlab='Age', 
            main='Distribution of patients\', who reported "Exemption 4", age',
            col=c(pastel_pie_color_vector[1], 
                  pastel_pie_color_vector[1], 
                  pastel_pie_color_vector[1], 
                  pastel_pie_color_vector[1], 
                  pastel_pie_color_vector[2], 
                  pastel_pie_color_vector[2],
                  pastel_pie_color_vector[2], 
                  pastel_pie_color_vector[2],
                  pastel_pie_color_vector[3],
                  pastel_pie_color_vector[3]))
```

Fig. 3. The histogram displaying distribution of age among patients who reported
"Exemption 4" age

\  

# Conclusions

To conclude, in the period of 2018 - 2020 only one product was considered as
concomitant to the causial of death, although it was reported along with other
products used at the time. The majority of death cases reported included 
"`r most_lethal_product_name_cap`", which is 
`r most_lethal_product_industry` product - no additional information about it
was provided, yet it might be responsible for around 95% of reported death 
cases.

# References

1. U.S. Food and Drug Administration. (2022, January 31). *CFSAN Adverse Event Reporting System (CAERS).*  https://www.fda.gov/food/compliance-enforcement-food/cfsan-adverse-event-reporting-system-caers.
2. U.S. Food and Drug Administration. (2020, December). *CAERS ASCII.* https://www.fda.gov/media/140575/download
3. U.S. Food and Drug Administration, Center for Food Safety and Applied Nutrition (CFSAN), Office of Analytics and Outreach (OAO). (2022, January). *README file for the Quaterly Data Extract (QDE) from the CFSAN Adverse Event Reporting System (CAERS).* https://www.fda.gov/media/97035/download
4. Huestegge, L., & Pötzsch, T. H. (2018). Integration processes during frequency graph comprehension: Performance and eye movements while processing tree maps versus pie charts. Applied Cognitive Psychology, 32(2), 200-216. https://doi.org/10.1002/acp.3396

